---
title: "Droughtnet Group 3"
authors: "Priscilla, Garrett & Emma"
output: html_document
date: "2023-01-23"
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#Code Outline
```{r}
# 1. Install and load packages
# 2. Import data and bind years (2020-2022)
# 3. Rename columns and filter dates
# 4. Check for NAs and misinputed data
# 5. Change column types
# 6. Importing corrected code file (DNCodes) and species glossary from Amy
# 7. Create a new joined table (Intermediate) that will have the new standardized information
# 8. More Data Sleuthing
 ## a. Transects (explained example)
 ## b. Hits 
 ## c. Block
 ## d. Plot
 ## e. Species Code
 ## f. Type
 ## g. Status
 ## h. Native Status
```
# 1. Install and Load Packages
```{r}
install.packages("tidyverse")
install.packages("readxl")
install.packages("janitor")

library(tidyverse)
library(readxl)
library(janitor)
library(dplyr)
```
# 2. Import data and bind years
```{r}
Spring_2020_raw <- read_excel("Spring 2020 - Percent Cover - Raw Data DNET.xlsx", sheet = "Sheet1", col_names = FALSE)
Spring_2021_raw <- read_excel("Spring 2021 - Percent Cover - Raw Data DNET.xlsx", sheet = "Sheet1", col_names = FALSE)
Spring_2022_raw <- read_excel("Spring 2022 - Percent Cover - Raw Data DNET.xlsx", 
    sheet = "Sheet1", col_names = FALSE)

Spring_Data_raw <- bind_rows(Spring_2020_raw, Spring_2021_raw, Spring_2022_raw)
```
# 3. Rename columns and filter dates
```{r}
Spring_Data_raw <- Spring_Data_raw %>% row_to_names(row_number(2)) %>% 
  rename(Transect = `Transect #`,  Hits = `# Hits`) %>% 
  filter(Date != "Date" & Date != "mm/dd/yy" & Date != "NA")
```
# 4. Check for NA's and misinputed Data. NA's fixed on Intermediate file
```{r}
unique(Spring_Data_raw$Date)
unique(Spring_Data_raw$Block)
unique(Spring_Data_raw$Plot)
unique(Spring_Data_raw$Treatment)

#Distance of 53 is a wrong input, changing to 3
Spring_Data_raw_better <- Spring_Data_raw %>% 
  mutate(Distance = ifelse(as.numeric(Distance) == 53, 3, as.numeric(Distance)))
unique(Spring_Data_raw$Distance)

```
# 5. Change Column Types
```{r}
Spring_Data_raw <- Spring_Data_raw %>%
  mutate(Block = as.numeric(Block)) %>% 
  mutate(Plot = as.numeric(Plot)) %>% 
  mutate(Transect = as.numeric(Transect))

Spring_Data_raw <- Spring_Data_raw %>% 
  mutate(Block = factor(Block)) %>% 
  mutate(Plot = factor(Plot)) %>% 
  mutate(Treatment = factor(Treatment)) %>% 
  mutate(Region = factor(Region)) %>% 
  mutate(Transect = as.numeric(Transect)) %>% 
  mutate(Distance = as.numeric(Distance)) %>%
  mutate(Hits = as.numeric(Hits)) %>% 
  mutate(Date = as.numeric(Date)) %>% 
  mutate(Date = as.Date(Date, origin = "1899-12-30"))
```
# 6. Importing Corrected Code file (DNCodes) and Species Glossary from Amy
```{r}
Full_Species_List_Glossary_reference_only_ <- read_excel("Full Species List Glossary (reference only).xlsx")
DNCodes <- read_csv("DNCodes.csv")

names(Full_Species_List_Glossary_reference_only_)

#Renaming columns
SpeciesKey <- Full_Species_List_Glossary_reference_only_ %>% 
  select(1:7) %>% 
  rename(SpeciesCode = 'Species Code (A-Z)',
         NativeStatus = 'Native/Non-Native',
         CommonName = 'Common Name',
         SciName = 'Full Scientific Name (Genus species)')
```
# 7. Create a new joined table (Intermediate) that will have the new standardized information
```{r}
Intermediate <- Spring_Data_raw %>%
  filter(is.na(Date) == FALSE) %>%
  full_join(DNCodes, by = "Code") %>% 
  left_join(SpeciesKey, by = "SpeciesCode") %>% 
  mutate(
    LifeForm = case_when(
          is.na(LifeForm) == FALSE  ~ Lifeform, 
          TRUE ~ as.character(LifeForm)),
    
    SpeciesCode = case_when(
      SpeciesCode == "L" ~ "litter",
      SpeciesCode == "B" ~ "bare ground",
      SpeciesCode == "W" ~ "coarse woody debris",
      SpeciesCode == "F" ~ "fine woody debris",
      SpeciesCode == "R" ~ "rock",
      SpeciesCode == "S" ~ "stem",
      SpeciesCode == "T" ~ "thatch",
      TRUE ~ as.character(SpeciesCode)
      ),
    
    SciName = case_when(
    Type == "Ground" ~ SpeciesCode,
    TRUE ~ as.character(SciName)
    ),
    
    CommonName = case_when(
      Type == "Ground" ~ SpeciesCode,
      TRUE ~ as.character(CommonName)
    )
  ) %>%
  select(-Code) %>%
  fill(Transect) %>%
  select(-Notes) %>%
  select(-LifeForm) %>% 
  mutate(Hits = coalesce(Hits,1)) %>%
  mutate(Status = ifelse(is.na(Status),"Ground",Status)) %>%
  mutate(NativeStatus = ifelse(is.na(NativeStatus), "Ground", NativeStatus))

Intermediate <- Intermediate %>%
  mutate(Transect = factor(Transect))
```
# 8. More Data Sleuthing
## a. Example for Transect Column
```{r}
#Find errors - identify unique values
Intermediate$Transect %>% unique()

#There are NAs in the Transect column. Find NAs (filter data for when NA is true).
Intermediate %>% filter(is.na(Transect) == TRUE)

#After looking back at the raw data for this instance, we determine that the NA is a result of the observer not recording the transect value. We can easily see that it is supposed to match the value of data row above it
#fill() will take the value from the data row above to populate the missing NA cell
#solution = line 113; fill(Transect)
```
# b. Data Sleuthing - Hits
```{R}
Intermediate$Hits %>% unique ()
Intermediate %>% filter(is.na(Hits) == TRUE)
#Hits - need to replace NAs to 1; because each species was hit at least once
#solution = line 116; mutate(Hits = coalesce(Hits,1))
```
# c. Data Sleuthing - Block
```{R}
Intermediate$Block %>% unique()
#no changes necessary
```
# d. Data Sleuthing - Plot
```{R}
Intermediate$Plot %>% unique()
#no changes necessary
```
# e. Data Sleuthing - SpeciesCode
```{R}
Intermediate$SpeciesCode %>% unique()
#missing a lot of species code data for 2020 and 2022 which will impact how Type, Status, SciName, Lifeform, Native Status, CommonName, old names, and Plant family will appear

#For 2022 - NAs due to entry error (did not enter speciescode) AND observer error (did not record ground cover for block 2 plot 8); easiest fix would probably revising the excel file itself

Intermediate[7097,10] <- UNK #replace NA found in species code in row 9097 with UNK
```
# f. Data Sleuthing - Type
```{R}
Intermediate$Type %>% unique()
Intermediate %>% filter(is.na(Type) == TRUE)
#lots of NAs, likely as a result of many NAs in the SpeciesCode column
```
# g. Data Sleuthing - Status
```{R}
Intermediate$Status %>% unique()
Intermediate %>% filter(is.na(Status) == TRUE)
#NAs are a result of groundcover categories not populating with a plant status, so changed all the NAs to ground
#solution=line 117; mutate(Status = ifelse(is.na(Status),"Ground",Status)
```
# h. Data Sleuthing - NativeStatus
```{R}
Intermediate$NativeStatus %>% unique()
Intermediate %>% filter(is.na(NativeStatus) == TRUE)
#NAs are a result of groundcover categories not populating with a plant status, so changed all the NAs to ground
#solution=line 118; mutate(Status = ifelse(is.na(Status),"Ground",Status)
```