---
title: "DNTransect_Summarizing"
author: "Amy Henry"
date: "2023-01-24"
output: html_document
---

```{r}
library(tidyverse)
library(vegan)
library(lubridate)

load(file = "DNTransect_Primary.Rdata")
```


Steps
Create index column
create one column per species using pivot wider
seperate metadata column from species data coulmns, keeping index in common
makes species data columns a matrix 

```{r}
#creating our index


DN_index <- DNTransect_Primary %>% 
  mutate(Index = paste(year(Date), Block, Plot, Transect, sep = "-"))

DN_Hits_Matrix <- DN_index %>% 
  filter(Type == "Plant") %>% 
  select(Index, SpeciesCode, Hits) %>%
  group_by(Index, SpeciesCode) %>% 
  summarize(Hits = sum(Hits)) %>% 
  pivot_wider(names_from = Index, values_from = Hits) %>% 
  replace(is.na(.), 0)
  
DN_PresAbs_Matrix <- DN_index %>% 
  filter(Type == "Plant") %>% 
  select(Index, SpeciesCode, Hits) %>% 
  group_by(Index, SpeciesCode) %>% 
  summarize(Hits = sum(Hits)) %>% 
  mutate(PresAbs = case_when(
    Hits >= 1 ~ 1,
    Hits < 1 ~ 0
  )) %>% 
   select(-Hits) %>% 
  pivot_wider(names_from = SpeciesCode, values_from = PresAbs) %>% 
  replace(is.na(.), 0)
  
DN_diversity <- DN_PresAbs_Matrix %>% 
  mutate(SpRich = rowSums(across(where(is.numeric)), na.rm=TRUE))
         


DN_diversity_index <- DN_diversity %>% 
  select(Index, SpRich) %>% 
  full_join(Transect_Index) %>%
  select(Index, Date:Transect, SpRich)


Transect_Index <- DN_index %>% 
  select(Date, Block:Transect, Index) %>%
  distinct()

DN_diversity_index %>% ggplot() +
  aes(x=Treatment, y=SpRich, fill = Region) +
  geom_boxplot()
```


```{r}
library(tidyverse)
Shannon <- DN_Hits_Matrix %>%  ungroup() %>% select(AVEFAT:SILSPP) %>% diversity(index = "shannon", MARGIN = 1)

Shannon2 <- DN_Hits_Matrix %>% select(Index) %>%  bind_cols(Shannon) %>% rename(H = '...2')

DN_diversity_index %>% full_join((Shannon2))

DN_diversity_index %>% ggplot() +
  aes(x = Treatment, y = H, fill = Region) + 
  geom_boxplot()
```


## Species Richness
Calculate species richness for each TRANSECT for each Treatment and Seed Source combination

-- Total
-- Native
-- Non-native
-- LifeForm
---Native
---Non-native
```{r}

```

## Shannon Diversity
Calculate Shannon Diversity for each TRANSECT for each Treatment and Seed Source combination

-- Total
-- Native
-- Non-native
-- LifeForm
---Native
---Non-native
```{r}

```


## Calculate Percent Cover with PRESENCE/ABSENCE
Convert Hits column to 0/1 (i.e., presence/absence), then sum for percent cover for each TRANSECT for each Treatment and Seed Source combination

-- For each Focal Species (ENCCAL, ERIFAS, ARTCAL, SALMEL)
-- Total
-- Native
-- Non-native
-- LifeForm
---Native
---Non-native

1. Convert our hits column in our original data to a 0/1 pres absence column
2. sum the observations per transect
3. divide by the total number of observations in the transect (9)
```{r}
SpeciesPerCover <- DN_index %>% 
  filter(Type == "Plant") %>% 
  select(Index, SpeciesCode, Hits) %>% 
  
  mutate(PresAbs = case_when(
    Hits >= 1 ~ 1, 
    Hits < 1 ~ 0
  )) %>% 

  group_by(Index, SpeciesCode) %>% 
  summarize(PresAbsSum = sum(PresAbs)) %>% 
  
  mutate(PerCover = PresAbsSum/9) %>% 
  select(-PresAbsSum) %>% 
  pivot_wider(names_from = SpeciesCode, values_from = PerCover) %>% 
  
  replace(is.na(.), 0)

```

```{r}
#Renaming columns
SpeciesKey <- Full_Species_List_Glossary_reference_only_ %>% 
  select(1:7) %>% 
  rename(SpeciesCode = 'Species Code (A-Z)',
         NativeStatus = 'Native/Non-Native',
         CommonName = 'Common Name',
         SciName = 'Full Scientific Name (Genus species)')

SpeciesPerCover %>% 
  pivot_longer(2:ncol(SpeciesPerCover), names_to = "SpeciesCode", values_to = "PerCover") %>% 
  left_join(Full_Species_List_Glossary_reference_only_, by)
  
```


## Calculate Rough Abundance with COUNTS OF HITS
Sum the count of hits for each TRANSECT for each Treatment and Seed Source combination

-- For each Focal Species (ENCCAL, ERIFAS, ARTCAL, SALMEL)
-- Total
-- Native
-- Non-native
-- LifeForm
---Native
---Non-native
```{r}

```



